\documentclass[12pt]{article}

\usepackage{listings}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{minted}

\lstset{
	frame = single,
	basicstyle=\ttfamily\footnotesize,
	breaklines=true
}

\title{Thesis report on dynamic instances}
\author{Albert ten Napel}
\date{}

\begin{document}
\maketitle

\section{Memoization}

Some needed utility functions and the State effect used.
\begin{minted}{haskell}
empty : NatMap
find : NatMap -> Nat -> Maybe Nat
update : NatMap -> Nat -> Nat -> NatMap

effect State {
  get : () -> NatMap
  set : NatMap -> ()
}

state : IntMap -> (Nat!{State} => Nat)
state s = handler {
  return v -> \s -> v
  get () k -> \s -> k s s
  set v k -> \_ -> k () v
  finally f -> f s 
}
\end{minted}

\newpage
\subsection{Basic algebraic effects}

Memoized fibonacci function.
\begin{minted}{haskell}
fib : Nat -> Nat!{State}
fib n =
  map <- get ();
  case find map n of
    Just m -> m
    Nothing ->
      let result =
        if n < 2 then
          n
        else
          fib (n - 1) + fib (n - 2);
      set (update map n result);
      result

result = with state empty handle
  fib 4 * fib 5
\end{minted}
Problem: the program cannot use State for other purposes since it's already used by the fibonacci.

\newpage
\subsection{Static instances}

We can solve the previously stated problem by using an instance of the State effects specifically used for the fibonacci function.
\begin{minted}{haskell}
instance State fibState

fib : Nat -> Nat!{fibState}
fib n =
  map <- fibState#get ();
  case find map n of
    Just m -> m
    Nothing ->
      let result =
        if n < 2 then
          n
        else
          fib (n - 1) + fib (n - 2);
      fibState#set (update map n result);
      result

result = with state fibState empty handle
  fib 4 * fib 5
\end{minted}

Problem: we cannot define a general memoization function since that would require generating an instance for each memoized function.

\newpage
\subsection{Dynamic instances}

Dynamic instances solve the previous problem by allowing us to dynamically generate an instance for each memoized function.
\begin{minted}{haskell}
memo : ((Nat -> Nat) -> Nat -> Nat) -> ((Nat -> Nat), Inst State)
memo f =
  let inst = new State;
  let rec n =
    map <- inst#get ();
    case find map n of
      Just m -> m
      Nothing ->
        let result = f rec n;
        inst#set (update map n result);
        result;
  (rec, inst)

fibR f n =
  if n < 2 then
    n
  else
    f (n - 1) + f (n - 2)
(fib, fibState) = memo fibR

facR f n =
  if n < 3 then
    n
  else
    n * f (n - 1)
(fac, facState) = memo facR

result =
  with state fibState empty handle
  with state facState empty handle
    fib 4 * fac 5
\end{minted}

\it{Can fibState or facState escape the handlers?}

\end{document}