\documentclass[12pt]{book}

\usepackage{xcolor}
\usepackage{mdframed}
%\usepackage{colorbox}
\usepackage{listings}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{adjustbox}
\usepackage{amsthm}
\usepackage{capt-of}
\usepackage{minted}
\usepackage{fitch}
\usepackage[parfill]{parskip}
\usepackage{empheq}
\usepackage{multicol}
\usepackage{mathpartir}
\usepackage{scalerel}

% infer macros
\usepackage{hyperref}
\usepackage{pftools}
\usepackage{locallabel}

\usepackage{enumitem}
\setlist{nolistsep}

\usepackage{cleveref}
\crefname{section}{§}{§§}
\Crefname{section}{§}{§§}
\crefformat{section}{§#2#1#3}

\setminted{
	breaklines=false,
	breakanywhere=false,
	breakbytoken=false,
}

\newtheorem{lemma}{Lemma}
\newtheorem{sublemma}{Lemma}[lemma]

\newtheorem{theorem}{Theorem}

\newcommand\ident[1]{\mintinline{haskell}{#1}}
\newcommand\identp[1]{\mintinline{text}{#1}}

\newcommand{\lang}{Miro}

\lstset{
	frame = single,
	basicstyle=\ttfamily\footnotesize,
	breaklines=false
}

\title{A type system for algebraic effects and handlers with dynamic instances}
\author{Albert ten Napel}
\date{}

\begin{document}
\maketitle

\tableofcontents
\newpage

\chapter{Introduction}
\input{introduction.tex}

\chapter[Algebraic effects introduction]{An introduction to algebraic effects and handlers}
\input{background.tex}

\chapter{Introduction to \lang{}}
\input{examples.tex}

\chapter[Algebraic effects theory]{Semantics and types of algebraic effects and handlers}
\input{background-theory.tex}

\chapter{Semantics and types of \lang{}}
\input{calculus.tex}

\chapter{Related work}
\input{related.tex}

\chapter{Conclusion and future work}
\input{conclusion.tex}
% \input{features.tex}

\newpage
\begin{thebibliography}{100}
\bibitem{eff1}
Bauer, Andrej, and Matija Pretnar. "Programming with algebraic effects and handlers." Journal of Logical and Algebraic Methods in Programming 84.1 (2015): 108-123.
\bibitem{eff2}
Bauer, Andrej, and Matija Pretnar. "An effect system for algebraic effects and handlers." International Conference on Algebra and Coalgebra in Computer Science. Springer, Berlin, Heidelberg, 2013.
\bibitem{links}
Hillerstr\"{o}m, Daniel, and Sam Lindley. "Liberating effects with rows and handlers." Proceedings of the 1st International Workshop on Type-Driven Development. ACM, 2016.
\bibitem{koka}
Leijen, Daan. "Type directed compilation of row-typed algebraic effects." POPL. 2017.
\bibitem{koka2}
Leijen, Daan. Algebraic Effects for Functional Programming. Technical Report. 15 pages. https://www. microsoft. com/en-us/research/publication/algebraic-effects-for-functional-programming, 2016.
\bibitem{kokadynamic}
Leijen, Daan. "First class dynamic effect handlers: or, polymorphic heaps with dynamic effect handlers." Proceedings of the 3rd ACM SIGPLAN International Workshop on Type-Driven Development. ACM, 2018.
\bibitem{kokaresources}
Leijen, Daan. Algebraic effect handlers with resources and deep finalization. Technical Report MSR-TR-2018-10. Microsoft Research, 2018.
%\bibitem{frank}
%Lindley, Sam, Conor McBride, and Craig McLaughlin. "Do Be Do. In: POPL'2017. ACM, New York, pp. 500-514. ISBN 9781450346603, http://dx. doi. org/10.1145/3009837.3009897."
%\bibitem{idris}
%Brady, Edwin. "Programming and Reasoning with Side-Effects in IDRIS." (2014).
\bibitem{fg-stlc}
Levy, PaulBlain, John Power, and Hayo Thielecke. "Modelling environments in call-by-value programming languages." Information and computation 185.2 (2003): 182-210.
\bibitem{anormalform}
Flanagan, Cormac, et al. "The essence of compiling with continuations." ACM Sigplan Notices. Vol. 28. No. 6. ACM, 1993.
\bibitem{handlewithcare}
Biernacki, Dariusz, et al. "Handle with care: relational interpretation of algebraic effects and handlers." Proceedings of the ACM on Programming Languages 2.POPL (2017): 8.
\bibitem{standardml}
Robin Milner, Mads Tofte, and David Macqueen. 1997. The Definition of Standard ML. MIT Press, Cambridge, MA, USA.ndar
\bibitem{haskell}
Jones, Simon Peyton, ed. Haskell 98 language and libraries: the revised report. Cambridge University Press, 2003.
\bibitem{algeff}
Plotkin, Gordon D., and Matija Pretnar. "Handling algebraic effects." arXiv preprint arXiv:1312.1399 (2013).
\bibitem{runst}
Launchbury, John, and Simon L. Peyton Jones. "Lazy functional state threads." ACM SIGPLAN Notices 29.6 (1994): 24-35.
\bibitem{typedlogic}
Claessen, Koen, and Peter Ljunglöf. "Typed Logical Variables in Haskell." Electr. Notes Theor. Comput. Sci. 41.1 (2000): 37.
%\bibitem{algtutorial}
%Pretnar, Matija. "An introduction to algebraic effects and handlers. invited tutorial paper." Electronic Notes in Theoretical Computer Science 319 (2015): 19-35.
\bibitem{handlewithcare}
Biernacki, Dariusz, et al. "Handle with care: relational interpretation of algebraic effects and handlers." Proceedings of the ACM on Programming Languages 2.POPL (2017): 8.
%\bibitem{abstractingalgeffects}
%Biernacki, Dariusz, et al. "Abstracting algebraic effects." Proceedings of the ACM on Programming Languages 3.POPL (2019): 6.
%\bibitem{tunneling}
%Zhang, Yizhou, and Andrew C. Myers. "Abstraction-safe effect handlers via tunneling." Proceedings of the ACM on Programming Languages 3.POPL (2019): 5.
\bibitem{effdirectly}
Kiselyov, Oleg, and K. C. Sivaramakrishnan. "Eff directly in OCaml." arXiv preprint arXiv:1812.11664 (2018).
\end{thebibliography}

\end{document}